---
title: "EUR LD panel held-out imputation comparison"
subtitle: "UKB EUR versus matched HGDP+1kGP EUR on five UKB-B regions"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
results <- read.csv("/data/opengwasdb/wip/eur-panel-imputation-comparison.csv")
results$panel <- factor(results$panel, levels = c("UKB EUR", "HGDP+1kGP EUR"))
results$region <- paste0("chr", results$chrom, ":", results$block)
eligible <- subset(results, n_truth >= 3)
successful <- subset(eligible, accepted == "True" & n_predicted >= 3 & is.finite(pearson_r))
```

## Question and result

This experiment asks a narrow practical question: if known UKB-B associations are
hidden, how accurately does reference completion recover their beta values using
the production UKB EUR panel versus the newly generated HGDP+1kGP EUR panel?

The primary estimand is the **within-analysis Pearson correlation between known
and imputed betas**, conditional on at least three held-out observations. Acceptance
rate is reported separately and is equally important: an LD panel should not look
better merely because the quality gate rejects difficult analyses.

**Recommendation: SHIP the AFR/EAS/SAS panels for ancestry-matched reference
completion.** The exploratory gate requires calibration acceptance of at least
98%, median held-out beta correlation of at least 0.95, and median paired loss
versus UKB EUR no greater than 0.02. HGDP+1kGP EUR passes at 98.9%, 0.963, and
0.014 respectively. This supports utility, not parity: UKB EUR remains the better
EUR panel and should stay in production.

```{r}
summarise_panel <- function(d) {
  e <- subset(eligible, panel == d)
  s <- subset(successful, panel == d)
  data.frame(
    panel = d,
    eligible.analysis.regions = nrow(e),
    accepted.pct = 100 * mean(e$accepted == "True" & e$n_predicted >= 3),
    median.Pearson.r = median(s$pearson_r),
    Q1.Pearson.r = unname(quantile(s$pearson_r, .25)),
    Q3.Pearson.r = unname(quantile(s$pearson_r, .75)),
    median.Spearman.r = median(s$spearman_r),
    median.RMSE = median(s$rmse),
    median.slope = median(s$slope)
  )
}
summary_table <- do.call(rbind, lapply(levels(results$panel), summarise_panel))
summary_table$median.RMSE <- formatC(summary_table$median.RMSE, format = "e", digits = 2)
knitr::kable(summary_table, digits = 3,
             caption = "Held-out beta recovery across five regions.")
```

```{r}
#| fig-cap: "Distribution of within-analysis held-out beta correlations. Only accepted analysis-region fits with at least three known held-out betas are shown; acceptance is tabulated separately above."
#| fig-width: 8
#| fig-height: 5
boxplot(pearson_r ~ panel, data = successful, ylim = c(-1, 1),
        ylab = "Pearson correlation: known vs imputed beta", xlab = "",
        col = c("#4472C4", "#ED7D31"), outline = FALSE)
abline(h = 0, col = "grey60", lty = 2)
```

## Design

- Source truth is `/data/opengwasdb/wip/ukb-b-c128.opengwasdb`, the observed-only
  store with 2,514 analyses. A beta is reconstructed as `z × se`.
- `/data/opengwasdb/wip/ukb-b-c128-completed.opengwasdb` establishes the production
  completion parameters and expected store lineage, but is **not** used as truth
  for originally missing cells: those values were already imputed.
- Five matched UKB block boundaries were selected across chromosomes 1, 6, 11,
  16, and 22. These are not hand-picked by outcome accuracy.
- Candidate targets must occur in the observed store and both LD panels. Within
  each region, 20% of these variants are selected once using seed `20260806` and
  hidden identically from both panels for every analysis.
- All other observed variants represented by a panel remain available for training.
  Thus target truth is paired, while each panel retains its real-world variant
  coverage.
- The production kernel is used: eigenvectors truncated at 90% cumulative
  variance, elastic-net regression, polynomial rescaling, minimum training
  correlation 0.7, and the ±1 Mb observed-z cap. Standard errors use the panel's
  allele frequency and the scalar-N model.

### Data-integrity correction

The first execution of this experiment was deliberately withheld from reporting:
it exposed an allele-orientation defect in the new panel constructor. PLINK's ALT
dosages had been decomposed while the output table declared the lexicographically
canonical effect allele, reversing the required LD sign whenever canonical EA was
REF. A signed-LD regression test now locks down this invariant, the constructor
orients centred dosages before SVD, and all generated ancestry panels were rebuilt.
The results here are from the corrected artifacts. This is also why held-out
validation is a release gate rather than only a descriptive benchmark.

```{r}
region_design <- unique(results[c("region", "n_shared_variants", "n_masked_variants")])
region_design <- region_design[order(region_design$region), ]
knitr::kable(region_design, row.names = FALSE,
             caption = "Shared candidate and masked variant counts.")
```

## Region-level results

```{r}
region_rows <- lapply(split(eligible, list(eligible$region, eligible$panel), drop = TRUE), function(e) {
  s <- subset(e, accepted == "True" & n_predicted >= 3 & is.finite(pearson_r))
  data.frame(region = e$region[1], panel = as.character(e$panel[1]),
             eligible = nrow(e), accepted.pct = 100 * nrow(s) / nrow(e),
             median.Pearson.r = if (nrow(s)) median(s$pearson_r) else NA,
             median.RMSE = if (nrow(s)) median(s$rmse) else NA)
})
region_table <- do.call(rbind, region_rows)
region_table <- region_table[order(region_table$region, region_table$panel), ]
region_table$median.RMSE <- formatC(region_table$median.RMSE, format = "e", digits = 2)
knitr::kable(region_table, row.names = FALSE, digits = 3,
             caption = "Accuracy and acceptance by region.")
```

## Paired panel difference

The most direct comparison retains only analysis-region pairs successfully imputed
by both panels and subtracts the HGDP+1kGP correlation from the UKB correlation.
This conditional contrast is accompanied by the unconditional acceptance rates
above.

```{r}
u <- subset(successful, panel == "UKB EUR", select = c(region, analysis_index, pearson_r))
h <- subset(successful, panel == "HGDP+1kGP EUR", select = c(region, analysis_index, pearson_r))
paired <- merge(u, h, by = c("region", "analysis_index"), suffixes = c(".ukb", ".hgdp"))
paired$difference <- paired$pearson_r.ukb - paired$pearson_r.hgdp
paired_summary <- data.frame(
  paired.analysis.regions = nrow(paired),
  median.UKB.minus.HGDP = median(paired$difference),
  mean.UKB.minus.HGDP = mean(paired$difference),
  UKB.higher.pct = 100 * mean(paired$difference > 0),
  equal.pct = 100 * mean(paired$difference == 0)
)
knitr::kable(paired_summary, digits = 3,
             caption = "Paired difference in held-out beta correlation.")
```

```{r}
#| fig-cap: "Paired difference in Pearson correlation; positive values favour UKB EUR."
#| fig-width: 8
#| fig-height: 5
hist(paired$difference, breaks = 60, col = "#70AD47", border = "white",
     xlab = "UKB EUR r − HGDP+1kGP EUR r", main = "Paired accuracy difference")
abline(v = 0, col = "black", lty = 2)
```

## Interpretation and limits

This is a calibration experiment, not a genome-wide equivalence claim. It covers
five blocks and UKB-B traits, whose participants are predominantly European and
therefore favour ancestry matching to the UKB panel. Correlations are unstable
when very few held-out betas are available, which is why fewer than three are
excluded. The fixed global mask tests variant recovery but does not reproduce every
possible missingness pattern. A release decision should add more blocks, repeated
masks, allele-frequency strata, and traits from non-UKB cohorts.

## Reproduction

```{bash}
#| eval: false
../opengwasdb/.venv/bin/python resources/scripts/ld-panel/compare_eur_imputation.py \
  --store /data/opengwasdb/wip/ukb-b-c128.opengwasdb \
  --ukb-panel /data/opengwasdb/reference/ukb-hg38 \
  --hgdp-panel /data/opengwasdb/reference/hgdp1kgp-hg38/panel \
  --out docs/data/eur-panel-imputation-comparison.csv \
  --mask-fraction 0.2 --seed 20260806 --thresh 0.9 --jobs 10

quarto render resources/scripts/eur-panel-imputation-comparison.qmd \
  --output-dir docs
```
