---
title: "Canonical trait mapping: pipeline, validation, and the first ukb-b round"
subtitle: "Mapping free-text GWAS trait labels to EFO terms with BioLORD retrieval and the Jev chooser (issue #161)"
date: 2026-09-28
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(jsonlite)
library(DT)

data_dir <- if (dir.exists("docs/data")) "docs/data" else "../../docs/data"
read_tsv_df <- function(name) {
  read.delim(file.path(data_dir, name), sep = "\t", quote = "",
             stringsAsFactors = FALSE, na.strings = "")
}
summary_json <- fromJSON(file.path(data_dir, "trait-mapping-summary.json"))
val <- read_tsv_df("trait-mapping-validation-ogs00011.tsv") |>
  mutate(correct = correct == "True", abstained = abstained == "True",
         hit = !is.na(source_rank))
ukb <- read_tsv_df("trait-mapping-ukb-b-round1.tsv")
adj <- read_tsv_df("trait-mapping-ogs00011-adjudication.tsv")
ukb_adj <- read_tsv_df("trait-mapping-ukb-b-sample-adjudication.tsv")

conf_t <- summary_json$production$round$thresholds$confidence
margin_t <- summary_json$production$round$thresholds$margin
pct <- function(x, digits = 1) sprintf(paste0("%.", digits, "f%%"), 100 * x)
comma <- function(x) format(x, big.mark = ",", scientific = FALSE, trim = TRUE)

confident <- function(d) d$confidence >= conf_t & d$margin >= margin_t
val_promoted <- val |> filter(!abstained, confident(val))
catalog_precision <- mean(val_promoted$correct)
defensible <- adj |> filter(verdict %in% c("pick more specific or more accurate", "equivalent"))
adjusted_precision <- (sum(val_promoted$correct) + nrow(defensible)) / nrow(val_promoted)
recall10 <- mean(val$hit)

bucket_n <- ukb |> count(bucket)
bn <- function(b) bucket_n$n[bucket_n$bucket == b]
auto <- ukb |> filter(bucket == "auto-accepted")
```

## Summary

The Canonical Trait Mapping Table maps free-text trait labels, such as UK
Biobank field descriptions, to ontology terms, so Analyses from different
sources can be joined on a shared trait. This report covers:

- the curation pipeline built for epic #161;
- its validation against the GWAS Catalog's own EFO curation (Store OGS-00011);
- the first production round over the ukb-b queue.

**Headline results**

- **Validation (OGS-00011, `r comma(nrow(val))` labels).**
  - Retrieval puts the Catalog's term in the top 10 for `r pct(recall10)` of labels.
  - At the promotion thresholds (confidence ≥ `r conf_t`, margin ≥ `r margin_t`), `r comma(nrow(val_promoted))` labels would be auto-accepted.
  - `r pct(catalog_precision)` of those agree exactly with the Catalog's term.
  - We reviewed the `r nrow(adj)` confident disagreements one by one. None is a clear chooser error: most choose a more specific or equivalent term. Counting those as correct, the estimated precision is **`r pct(adjusted_precision)`**.
- **Production (ukb-b, `r comma(nrow(ukb))` labels).**
  - `r bn("auto-accepted")` rows were auto-accepted; they resolve `r comma(sum(auto$analyses))` Analyses (PR #180).
  - `r comma(bn("review (uncertain pick)") + bn("review (uncertain abstention)"))` labels go to human review.
  - `r comma(bn("no suitable term"))` labels were confidently abstained: no candidate their shortlist offered denoted the trait. Many of them genuinely have no EFO term (individual foods, keratometry, many questionnaire items), but some are retrieval misses — a valid EFO term exists and was simply not shortlisted (e.g. the ICD-10 labels C61 prostate cancer, K59.0 constipation, and H00.1 chalazion), so they may become mappable once retrieval improves.
  - In a random sample of 60 auto-accepted rows, `r sum(ukb_adj$verdict == "matches the label")` match the label exactly. The other `r sum(ukb_adj$verdict != "matches the label")` use a broader term where EFO has nothing more specific. None is wrong.
- **Cost.** Both rounds together cost **$`r sprintf("%.2f", summary_json$validation$cost_usd + summary_json$production$cost_usd)`**: `r comma(summary_json$validation$input_tokens + summary_json$production$input_tokens)` input tokens over `r comma(nrow(val) + nrow(ukb))` per-trait requests.

::: {.callout-important}
## Issues found only by running on real data

Several defects passed a hermetic test suite of more than 500 tests. They
showed up only against the live Jev API, the real EFO release, or the real
embedding server. These classes of failure need live smoke tests. They are
listed in [Issues found by live runs](#issues).
:::

## The pipeline

```{mermaid}
flowchart LR
  M[Release Manifests<br/>analyses.tsv] --> G[gap-scan<br/>unmapped labels]
  O[EFO v3.94.0 OBO] --> I[ontology index<br/>canonical CURIEs]
  I --> EO[embed-ontology<br/>BioLORD-2023, once]
  G --> ET[embed-traits<br/>per round]
  I --> C[candidates<br/>lexical + semantic, RRF, top 10]
  EO --> C
  ET --> C
  C --> CH[choose<br/>one Jev request per trait]
  CH --> R[reduce<br/>proposals, ledger, buckets]
  R --> P[promote<br/>thresholds gate]
  P --> T[(mapping.tsv)]
  P --> Q[review queue]
  P --> N[no-suitable-term]
```

Each stage is an idempotent `pixi run -e curation round-*` task. It reads
its pins (ontology release, embedding model and store `build_id`, chooser
model, thresholds) from `<round-dir>/round.yaml`, and it refuses to run when
an artifact contradicts a pin.

| Stage | What it does |
|---|---|
| `ontology-index` | Parses the EFO OBO and normalises every id to a canonical CURIE (`efo:EFO_0003939` → `EFO:0003939`, OBO PURLs → `PREFIX:local`). It drops `r comma(summary_json$index$excluded_non_curie)` gene and dbpedia IRIs, leaving `r comma(summary_json$index$terms)` terms. |
| `embed-ontology` | Embeds each term's label, synonyms and definition with BioLORD-2023, served by Text Embeddings Inference. The result is a resumable numpy store, content-addressed by `build_id`. |
| `gap-scan` | Queues every distinct normalised trait label that is not already in the mapping table. |
| `embed-traits` | Embeds the queued labels with the same model. |
| `candidates` | Runs four lexical channels (exact, normalised, token overlap, synonym) and cosine nearest neighbours, fuses them by reciprocal rank, and keeps the top 10. An obsolete term is replaced by its `replaced_by` successor, or dropped if it has none. |
| `choose` | Sends one Jev request per trait, with `none_suitable` always offered. Each result is written atomically to its own YAML file, keyed by a request fingerprint. A rerun skips traits whose fingerprint is unchanged. A cumulative cost cap (`--max-cost-usd`) bounds spend across reruns. |
| `reduce` | Aggregates the per-trait results into proposals, a cost ledger, and a bucket reconciliation. The buckets must sum to the queue. |
| `promote` | Appends a proposal only if confidence ≥ `r conf_t` and runner-up margin ≥ `r margin_t`. Uncertain picks and uncertain abstentions go to the review queue. Confident abstentions are recorded as unmapped by design. `none_suitable` never reaches the table. |

### Design decisions

**One request per trait, not packed batches.** TypeSafe has no asynchronous
or discounted batch API. We compared packing 19 traits into one request with
sending them individually:

- Packing saved about 20% of input tokens: roughly 3 cents over the whole ukb-b queue.
- Packing changed the answers:
  - single and packed requests disagreed on 1 of 19 choices;
  - two packed orderings of the same traits disagreed on 2 of 19;
  - confidence moved by up to 0.39 depending on which other traits shared the request.

We chose per-trait requests: each answer depends only on its own shortlist,
can be cached, and can be retried on its own.

**Always offer `none_suitable`.** Most ukb-b labels have no EFO term. Without
an explicit way to abstain, the chooser has to pick the nearest related term.
In the validation run, when the Catalog's term was *not* in the shortlist, Jev
abstained on `r pct(mean(val$abstained[!val$hit]))` of labels rather than
picking a wrong term.

**Thresholds taken from validation, not guessed.** The defaults (0.85 / 0.20)
were kept after calibration (below). Above confidence 0.8 the margin
threshold is redundant: a confident pick has a large margin by construction.

## Issues found by live runs {#issues}

Every row below had passing unit tests at the time. Each one is now fixed and
covered by a regression test.

| Issue | Found by | Impact if unfixed |
|---|---|---|
| Jev returns probabilities rounded to 2 decimal places, so a 21-option distribution summed to 0.99 and failed a 1e-6 check | Live Jev pilot, 25 labels | 3 of 25 paid answers were thrown away |
| EFO v3.94.0 spells native ids as `efo:EFO_0003939`; every manifest uses `EFO:0003939` | Live pilot shortlist | No mapping would ever match a source id, and every promoted row would have had a non-canonical id |
| 12% of shortlist rows (3,055 of 25,020) were obsolete terms | ukb-b shortlist inspection | Obsolete terms could be chosen and written to the table |
| TEI accepts at most 32 inputs per request; the client sent 128 | Live TEI server | Every embedding request would fail |
| 6 imported terms have no label, synonym, or definition; TEI refuses empty input | Live TEI server | Ontology embedding aborted |
| The client posted to the base URL, not `/v1/embeddings` | Live TEI server | Every embedding request would fail |
| 332 GWAS Catalog rows map one trait to several terms (a comma-joined IRI list) | OGS-00011 harvest | Each would score as a certain error, biasing calibration down |
| `choose` and `reduce` computed different fingerprints for a Jev round | Code review | A real round could never complete |
| The cost cap reset on every rerun and was overshot by requests in flight | Code review | Unbounded spend across resumed runs |
| A stub or fixture round could promote into the tracked mapping table (the default resource) | A reviewer running the runbook | Test rows written to `mapping.tsv` and the version bumped |
| A curator `amend` could write `none_suitable` into the table | Code review | The sentinel stored as a term |

## Validation on OGS-00011 {#validation}

OGS-00011 is the GWAS Catalog EUR hybrid Store. Almost every Analysis carries
a curator-assigned EFO term (`source_provided`). We harvested its
single-term mappings, excluding 332 multi-term rows. That left
`r comma(nrow(val))` distinct labels, which we ran through exactly the same
round as production.

The Catalog's curation is a **silver standard**: it is often deliberately
broad (e.g. *pulmonary function measurement* for FEV1). A pick that differs
from it is not necessarily wrong. That is why confident disagreements were
reviewed individually below.

### Retrieval

```{r retrieval}
recall_at <- tibble(k = 1:10) |>
  rowwise() |>
  mutate(recall = mean(!is.na(val$source_rank) & val$source_rank <= k)) |>
  ungroup()
by_stratum <- val |> group_by(stratum) |>
  summarise(labels = n(), recall_at_10 = mean(hit), .groups = "drop")

ggplot(recall_at, aes(k, recall)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = "shortlist size k", y = "Catalog term within top k",
       title = "Retrieval recall@k (lexical + BioLORD, reciprocal rank fusion)") +
  theme_minimal()
knitr::kable(by_stratum |> mutate(recall_at_10 = pct(recall_at_10)),
             col.names = c("Stratum", "Labels", "Recall@10"))
```

Retrieval sets the ceiling on accuracy: the chooser can only pick from the
shortlist. The Catalog's exact term is missing for `r pct(1 - recall10)` of
labels. Often a close alternative is present instead, as the adjudication
shows.

### Choice

```{r choice}
choice_tab <- val |>
  mutate(retrieval = if_else(hit, "Catalog term shortlisted", "Catalog term not shortlisted"),
         outcome = case_when(abstained ~ "abstained (none_suitable)",
                             correct ~ "picked the Catalog term",
                             TRUE ~ "picked another term")) |>
  count(retrieval, outcome) |>
  group_by(retrieval) |> mutate(share = pct(n / sum(n))) |> ungroup()
knitr::kable(choice_tab, col.names = c("Retrieval", "Jev outcome", "Labels", "Share"))
```

### Thresholds: precision against yield

```{r thresholds, fig.height=4}
grid <- expand_grid(conf = seq(0.5, 0.95, by = 0.05)) |>
  rowwise() |>
  mutate(promoted = sum(!val$abstained & val$confidence >= conf & val$margin >= margin_t),
         precision = mean(val$correct[!val$abstained & val$confidence >= conf & val$margin >= margin_t]),
         yield = sum(val$correct & !val$abstained & val$confidence >= conf & val$margin >= margin_t) / nrow(val)) |>
  ungroup()
grid |>
  rename(`agreement with Catalog` = precision) |>
  pivot_longer(c(`agreement with Catalog`, yield)) |>
  ggplot(aes(conf, value, colour = name)) +
  geom_line() + geom_point() +
  geom_vline(xintercept = conf_t, linetype = "dashed") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = "confidence threshold (margin ≥ 0.2)", y = NULL, colour = NULL,
       title = "Exact agreement with the Catalog term, and yield (correct / all labels)") +
  theme_minimal()
knitr::kable(grid |> mutate(precision = pct(precision), yield = pct(yield)),
             col.names = c("Confidence ≥", "Promoted", "Agreement", "Yield"))
```

```{r reliability, fig.height=3.5}
val |> filter(!abstained) |>
  mutate(bin = cut(confidence, breaks = seq(0.3, 1, by = 0.1), include.lowest = TRUE)) |>
  group_by(bin) |> summarise(n = n(), accuracy = mean(correct), .groups = "drop") |>
  ggplot(aes(bin, accuracy)) +
  geom_col(fill = "grey60") +
  geom_text(aes(label = paste0("n=", n)), vjust = -0.3, size = 3) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = "Jev confidence", y = "agreement with Catalog",
       title = "Reliability: agreement rises with confidence") +
  theme_minimal()
```

### Adjudicating the confident disagreements

We reviewed all `r nrow(adj)` promoted-but-disagreeing picks. The Catalog's
term was sometimes itself wrong: "nonsyndromic cleft palate" is mapped to
*cleft lip*, and "collagen band thickness" to *blood protein amount*.

::: {.callout-note}
These verdicts were made by the AI assistant (Claude) that built the
pipeline, by reading labels and term names. They are not a human curator's
review, and they are published here in full so they can be checked.
:::

```{r adjudication}
knitr::kable(adj |> count(verdict, sort = TRUE), col.names = c("Verdict", "Labels"))
datatable(adj |> select(trait_label, source_label, selected_label, confidence, verdict, note),
          colnames = c("Trait label", "Catalog term", "Jev pick", "Conf.", "Verdict", "Note"),
          rownames = FALSE, options = list(pageLength = 10, scrollX = TRUE))
```

Counting *more specific or more accurate* and *equivalent* picks as correct,
the estimated precision at the production thresholds is
**`r pct(adjusted_precision)`**, compared with `r pct(catalog_precision)`
agreement with the Catalog's own term.

## ukb-b round 1 {#ukb-b}

The ukb-b queue has `r comma(nrow(ukb))` distinct labels, covering
`r comma(sum(ukb$analyses))` Analyses. None was in the mapping table
beforehand. The round cost $`r sprintf("%.4f", summary_json$production$cost_usd)`,
with no failed requests.

```{r ukb-buckets}
knitr::kable(ukb |> group_by(bucket) |>
               summarise(labels = n(), analyses = sum(analyses), .groups = "drop") |>
               arrange(desc(labels)),
             col.names = c("Outcome", "Labels", "Analyses"))
```

```{r ukb-fields, fig.height=4.5}
ukb |>
  mutate(bucket = factor(bucket, levels = c("auto-accepted", "review (uncertain pick)",
                                            "review (uncertain abstention)", "no suitable term"))) |>
  count(field, bucket) |>
  ggplot(aes(reorder(field, n, sum), n, fill = bucket)) +
  geom_col() + coord_flip() +
  scale_fill_manual(values = c("#2a9d8f", "#e9c46a", "#f4a261", "#bcbcbc")) +
  labs(x = NULL, y = "labels", fill = NULL, title = "Outcome by UK Biobank field family") +
  guides(fill = guide_legend(nrow = 2)) +
  theme_minimal() + theme(legend.position = "bottom")
knitr::kable(ukb |> count(field, bucket) |>
               pivot_wider(names_from = bucket, values_from = n, values_fill = 0) |>
               mutate(total = rowSums(across(where(is.numeric)))) |>
               arrange(desc(total)) |>
               select(field, total, `auto-accepted`, `review (uncertain pick)`,
                      `review (uncertain abstention)`, `no suitable term`),
             col.names = c("Field family", "Labels", "Auto-accepted", "Review: pick",
                           "Review: abstention", "No suitable term"))
admin <- ukb |> filter(field == "Hospital administrative field")
```

Clinical codes map well: ICD-10 diagnoses, self-reported illnesses and
symptoms, and family history. Diet items ("beef intake", "dried fruit
intake") and keratometry fields mostly have no EFO term, so abstention is the
right answer for them. But not every abstention is a missing term: the
ICD-10-coded labels `diagnoses - main icd10: c61 malignant neoplasm of
prostate`, `... k59.0 constipation`, and `... h00.1 chalazion` were abstained
because no candidate was retrieved, even though EFO has terms for prostate
cancer, constipation, and chalazion — their ICD-10 xrefs exist, but the
label's full text did not shortlist them at retrieval time, so the chooser
never saw them. Medications are rarely auto-accepted: EFO has few
drug-exposure terms, so they go to review.

::: {.callout-warning}
## `r nrow(admin)` ukb-b "traits" are hospital administrative fields

For example: "PCT where patient's GP was registered", "main speciality of
consultant", and hospital admission or discharge methods. These are not
phenotypes. Jev correctly declined `r sum(admin$bucket == "no suitable term")`
of them outright. Only one was auto-accepted: "methods of discharge: patient
death" → *dead (follow-up status)*, which is defensible. These fields should
arguably be excluded from the mapping queue, or from trait-level use, upstream.
:::

### Checking a sample of auto-accepted rows

We drew a random sample of 60 of the `r nrow(auto)` auto-accepted rows and
reviewed them the same way (AI assistant review, not a human curator):
`r sum(ukb_adj$verdict == "matches the label")` match the label, and
`r sum(ukb_adj$verdict != "matches the label")` use a broader term where EFO
has nothing more specific. For example, "qualifications: O levels" maps to
*educational attainment*, and "birth weight of first child" maps to *birth
weight* (EFO has no offspring-birth-weight term). None was wrong.

```{r ukb-sample}
datatable(ukb_adj |> select(trait_label, selected_id, selected_label, confidence, verdict, note),
          colnames = c("Trait label", "Term", "Label", "Conf.", "Verdict", "Note"),
          rownames = FALSE, options = list(pageLength = 10, scrollX = TRUE))
```

### All ukb-b outcomes

```{r ukb-all}
datatable(ukb |> select(trait_label, field, analyses, selected_id, selected_label, confidence, bucket),
          colnames = c("Trait label", "Field", "Analyses", "Term", "Label", "Conf.", "Outcome"),
          rownames = FALSE, filter = "top",
          options = list(pageLength = 15, scrollX = TRUE))
```

## Limitations and next steps

- **The review queue is unreviewed.** `r comma(bn("review (uncertain pick)") + bn("review (uncertain abstention)"))` ukb-b labels are waiting for a human curator. The reviewed-queue round trip (accept, amend, reject) is implemented but has not been used on real data yet.
- **Adjudication was done by AI.** The disagreement and sample verdicts are the assistant's, not a curator's. A human spot check of the auto-accepted rows in PR #180 should come before relying on them widely.
- **Retrieval is the ceiling.** Recall@10 is `r pct(recall10)`. A larger shortlist would raise it: Jev accepts up to 254 options, and cost grows roughly linearly with shortlist size. Measure this before the next round.
- **The Catalog is a silver standard.** Exact agreement understates precision. Some Catalog mappings are coarser than the label, and a few are simply wrong.
- **Lower thresholds may be viable.** Agreement at confidence ≥ 0.7 is `r pct(grid$precision[grid$conf == 0.7])` against the Catalog, before adjudication. Adjudicating that band would show whether the threshold can drop and raise yield.
- **The `none_suitable` list is not all missing terms.** It mixes labels EFO genuinely has no term for (individual foods, keratometry fields, hospital administrative fields) with retrieval misses — labels whose term exists in EFO but was not shortlisted, such as the ICD-10 codes C61 (prostate cancer), K59.0 (constipation), and H00.1 (chalazion); see the ukb-b section. Mapping the first group needs new EFO terms (an upstream request) or a non-EFO target; the second group may map once #185 lands, so a better chooser is not ruled out.

## Reproducing

```sh
# once per EFO release (TEI serving BioLORD-2023 on localhost:8080)
pixi run -e curation ontology-index --obo efo.obo --release efo/v3.94.0 \
    --output .cache/curation/efo-v3.94.0.index.json
pixi run -e curation embed-ontology --ontology-index .cache/curation/efo-v3.94.0.index.json \
    --output .cache/curation/efo-v3.94.0--BioLORD-2023.ontology-embeddings

# a round (production: --manifests; validation: --queue-tsv from `harvest`)
pixi run -e curation round-init --round-dir <dir> --index <index> \
    --ontology-embeddings <store> --chooser jev --manifests <analyses.tsv> \
    --threshold-evidence "<calibration>"
for step in gap-scan embed-traits candidates; do pixi run -e curation round-$step --round-dir <dir>; done
pixi run -e curation round-choose --round-dir <dir> --workers 8 --max-cost-usd 1
pixi run -e curation round-reduce --round-dir <dir>
pixi run -e curation round-promote --round-dir <dir>

# this report's data, then the page
pixi run -e curation python resources/scripts/export-trait-mapping-report.py \
    --index <index> --validation <ogs-00011.tsv> \
    --validation-round <validation dir> --production-round <ukb-b dir>
pixi run --environment docs quarto render resources/scripts/trait-mapping-report.qmd --to html
```

Pins for both rounds: EFO `r summary_json$production$round$ontology_release`;
`r summary_json$production$round$embedding_model` (ontology store
`r summary_json$production$round$embedding_build_id`); Jev
`r summary_json$production$round$chooser$model`; shortlist size
`r summary_json$production$round$shortlist_size`.
